---
title: "pollution_time_series"
author: "Yuwei Hu"
date: "4/30/2022"
output:
  html_document: default
  pdf_document: default
---

```{r, include=FALSE}
library(xts)
library(tidyverse)
library(dygraphs)
library(forecast)
library(ggplot2)
library(ggpubr)
```


## Data
```{r,echo = FALSE}
data <- read.csv("data/cleaned_pollution_data.csv")
data <- data[,-1]
data1 <- data%>%
  select(SITE_ID,STATE,COUNTY,TNH4,K,TOTAL_NO3,date)%>%
  filter(STATE == "NY")

data_month <- data1 %>% 
  mutate(month_year = paste(format(as.Date(date), "%Y-%m"),'01', sep='-')) %>% 
  group_by(month_year) %>% 
  summarise(aver_TNH4 = mean(TNH4), aver_K = mean(K), aver_NO3 = mean(TOTAL_NO3)) %>% 
  ungroup()
```

## Visualization
```{r,echo  = FALSE}
g1<-ggplot(data_month, aes(as.Date(month_year), aver_TNH4)) + 
  geom_line()+
  labs(x = "", y = "TNH4")+
  ggtitle("Time series of TNH4")

g2<-ggplot(data_month, aes(as.Date(month_year), aver_K)) + 
  geom_line()+
  labs(x = "", y = "K")+
  ggtitle("Time series of K")

g3<-ggplot(data_month, aes(as.Date(month_year), aver_NO3)) + 
  geom_line()+
  labs(x = "", y = "NO3")+
  ggtitle("Time series of NO3")

ggarrange(g1,g2,g3, nrow = 2, ncol = 2)
```


## Stationarity

From the time series graphs, we observed that time series of K is stationary, but time series of TNH4 and NO3 are not, which appear decreasing trend. Thus, we decompose time series of TNH4 and NO3 to separate them into trend and other components.

**Decomposition of TNH4 data**

```{r,echo = FALSE}
data_month<-data_month%>%
  mutate(year = format(as.Date(month_year), "%Y"),month = format(as.Date(month_year), "%b"))
TNH4series<-ts(data_month$aver_TNH4,start=c(1999,1), end=c(2019,1), frequency=12)
TNH4decomp<-decompose(TNH4series) 
plot(TNH4decomp) 
```

**Decomposition of NO3 data**

```{r,echo = FALSE}
NO3series<-ts(data_month$aver_NO3,start=c(1999,1), end=c(2019,1), frequency=12)
NO3decomp<-decompose(NO3series) 
plot(NO3decomp) 
```

We observed that the trend in decomposition of these two time series is decreasing, same as what we supposed above.

## Differencing

```{r,echo = FALSE}
tseries_diff_TNH4 <- diff(TNH4series)
plot(tseries_diff_TNH4, ylab = "TNH4 (diff)", main = "Differencing TNH4 time series",)
```

```{r,echo = FALSE}
tseries_diff_NO3 <- diff(NO3series)
plot(tseries_diff_NO3, ylab = "NO3 (diff)", main = "Differencing NO3 time series")
```


## ACF and PACF

```{r,echo = FALSE}
par(mfrow=c(1,2))
acf(tseries_diff_TNH4)
pacf(tseries_diff_TNH4)
```

In ACF and PACF chart of TNH4 time series, we observe that ACF cuts off at 3rd point and PACF cuts off at 2nd point.

```{r,echo = FALSE}
Kseries<-ts(data_month$aver_K,start=c(1999,1), end=c(2019,1), frequency=12)
par(mfrow=c(1,2))
acf(Kseries)
pacf(Kseries)
```

In ACF and PACF chart of K time series, we observe that ACF cuts off at 4th point and PACF cuts off at 1st point.

```{r,echo = FALSE}
par(mfrow=c(1,2))
acf(tseries_diff_NO3)
pacf(tseries_diff_NO3)
```

In ACF and PACF chart of NO3 time series, we observe that ACF cuts off at 3th point and PACF cuts off at 2st point.

## Fitting in ARIMA model

```{r,echo = FALSE}
fit_TNH4 <- arima(TNH4series, order=c(1,1,2))
fit_TNH4
```

We fit a ARIMA model for TNH4 time series including ARIMA(1,1,2) and ARIMA(1,1,3). ARIMA(1,1,2) has a smaller AIC value.

```{r,echo = FALSE}
fit_K <- arima(Kseries, order=c(4,1,2))
fit_K
```

We fit a ARIMA model for K time series including ARIMA(4,1,1) and ARIMA(4,1,2). ARIMA(4,1,2) has a smaller AIC value.

```{r,echo = FALSE}
fit_NO3 <- arima(NO3series, order=c(2,1,4))
fit_NO3
```

We fit a ARIMA model for NO3 time series including ARIMA(2,1,3) and ARIMA(2,1,4). ARIMA(2,1,4) has a smaller AIC value.

## Forecasting by ARIMA
```{r,echo = FALSE}
autoplot(forecast(fit_TNH4),ylab = "TNH4", main = "Forecast for TNH4")
forecast(fit_TNH4,level = c(95))
```

```{r,echo = FALSE}
autoplot(forecast(fit_K),ylab = "K", main = "Forecast for K")
forecast(fit_K,level = c(95))
```

```{r,echo = FALSE}
autoplot(forecast(fit_NO3),ylab = "NO3", main = "Forecast for NO3")
forecast(fit_NO3,level = c(95))
```